Concurrent Programming with CXL

### 2020-06-14 book/paper.tex
Latest
Added the -m flag

To see that invariant $I_1$ is maintained, note that
\texttt{\^{}p == True} upon entry of \texttt{tas}.
So there are two cases:
\begin{enumerate}
\item \texttt{\^{}s} is \texttt{False} upon entry to \texttt{tas}.
Then upon exit \texttt{\^{}p == False} and \texttt{\^{}s == True}, maintaining
the invariant.
\item \texttt{\^{}s} is \texttt{True} upon entry to \texttt{tas}.
Then upon exit nothing has changed, maintaining the invariant.
\end{enumerate}
Invariant $I_1$ is also easy to verify for exiting the critical section.
Invariant $I_2(i)$ is obvious as (i) process $i$ only proceeds to the critical
section if \texttt{private[$i$] == False}, and (ii) no other process modifies
\texttt{private[$i$]}.

@@

\begin{figure}
\begin{code}
\begin{verbatim}
def tas(lk):
    atomic:
        result = ^lk;
        ^lk = True;
    ;
;
def lock(lk):
    while tas(lk):
        pass;
    ;
;
def unlock(lk):
    ^lk = False;
;
def Lock():
    result = False;
;
\end{verbatim}
\end{code}
\caption{The \texttt{Lock} interface in the \texttt{synch} module.}
\label{fig:locks}
\end{figure}

@@

Locks are probably the most popular and basic form of synchronization
in concurrent programs.  For this reason, CXL has a module called
\texttt{synch} that includes support for locks.
Figure~\ref{fig:locks} shows how they are implemented, and
Figure~\ref{fig:incfixed} gives an example of how they may be used,
in this case to fix the program of Figure~\ref{fig:inc}.
Notice that the module completely hides the implementation of the
lock.
The \texttt{synch} module includes a variety of other useful
synchronization primitives, which will be discussed in later
chapters.

@@

The astute reader might wonder if obtaining an ordinary lock itself is an
example of busy-waiting.  After all, the CXL \texttt{synch} implementation
of \texttt{lock()} spins in a loop until the lock is available
(see Figure~\ref{fig:locks}).
In most operating systems and programming language runtimes, however,
when a process acquires a lock, the process is placed on a scheduling
queue and stops using CPU cycles until the lock becomes available.
Moreover, if there are multiple processes waiting for a lock, there
is a scheduler that decides which of the processes gets the lock.
Busy waiting disables all this.

@@

We call a process \emph{blocked} in a state if steps by that process
do not change the shared memory in that state.  A process trying to
acquire a test-and-set spinlock held by another process is a good example
of a process being blocked: the process only reads the shared memory
but does not change it.

However, a process trying to obtain a read or write lock in
Figure~\ref{fig:rwbusy} has to actually obtain a spinlock just to check
if the read or write lock is available.  So even when the read or write
lock is not available, the process ends up updating the spinlock and
therefore the shared memory.

@@

\chapter{Starvation}

>>>>

$I_1$ is an inductive invariant.
To see that invariant $I_1$ is maintained, note that
\texttt{\^{}p == True} upon entry of \texttt{tas}.
So there are two cases:
\begin{enumerate}
\item \texttt{\^{}s} is \texttt{False} upon entry to \texttt{tas}.
Then upon exit \texttt{\^{}p == False} and \texttt{\^{}s == True}, maintaining
the invariant.
\item \texttt{\^{}s} is \texttt{True} upon entry to \texttt{tas}.
Then upon exit nothing has changed, maintaining the invariant.
\end{enumerate}
Invariant $I_1$ is also easy to verify for exiting the critical section.
Invariant $I_2(i)$ is obvious as (i) process $i$ only proceeds to the critical
section if \texttt{private[$i$] == False}, and (ii) no other process modifies
\texttt{private[$i$]}.

@@

\begin{figure}
\begin{code}
\begin{verbatim}
def tas(lk):
    atomic:
        result = ^lk;
        ^lk = True;
    ;
;
def Lock():
    result = False;
;
def lock(lk):
    while tas(lk):
        pass;
    ;
;
def unlock(lk):
    ^lk = False;
;
\end{verbatim}
\end{code}
\caption{The \texttt{Lock} interface in the \texttt{synch} module.}
\label{fig:spinlocks}
\end{figure}

@@

Locks are probably the most popular and basic form of synchronization
in concurrent programs.  For this reason, CXL has a module called
\texttt{synch} that includes support for locks.
Figure~\ref{fig:spinlocks} shows how they are implemented, and
Figure~\ref{fig:incfixed} gives an example of how they may be used,
in this case to fix the program of Figure~\ref{fig:inc}.
Notice that the module completely hides the implementation of the
lock.
The \texttt{synch} module includes a variety of other useful
synchronization primitives, which will be discussed in later
chapters.

We call a process \emph{blocked} in a state if the process is in
an \emph{infinite loop} in which the process does not update the
shared memory.
A process trying to
acquire a test-and-set spinlock held by another process is a good example
of a process being blocked: the process only reads the shared memory
but does not change it.
The only way to break out of the infinite loop is if another process
changes the shared memory.
The \texttt{while} loop in Peterson's algorithm, in case its condition
does not hold, is another example of a process being blocked.

\begin{figure}
\begin{code}
\begin{verbatim}
import list;

def Lock():
    result = dict{ .locked: False, .suspended: [] };
;
def lock(lk):
    atomic:
        if (^lk).locked:
            call stop (^lk).suspended;
            assert (^lk).locked;
        else:
            (^lk).locked = True;
        ;
    ;
;
def unlock(lk):
    atomic:
        if (^lk).suspended == []:
            (^lk).locked = False;
        else:
            go (head((^lk).suspended)) ();
            (^lk).suspended = tail((^lk).suspended);
        ;
    ;
;
\end{verbatim}
\end{code}
\caption{The \texttt{Lock} interface using suspension.}
\label{fig:suspension}
\end{figure}

In most operating systems, processes are virtual and can be suspended
until some condition changes.  For example, a process can be suspended
until the lock is available to it.
In CXL, a process can suspend itself and save its context in a
list.  Recall that the context of a process consists of its name tag,
its program counter, and the contents of its register and stack.
A context is a regular CXL value.
The syntax of the expression is as follows:

\begin{code}
\begin{verbatim}
stop L
\end{verbatim}
\end{code}

Here \texttt{L} is a shared list.
Another process can revive the process using the \texttt{go} stateent:

\begin{code}
\begin{verbatim}
go C R
\end{verbatim}
\end{code}

Here \texttt{C} is a context and \texttt{R} is a CXL value.
It causes a process with context \texttt{C} to be added to the state that has
just executed the \texttt{stop} expression.  The \texttt{stop} expression
returns the value \texttt{R}.

There is a second version of the \texttt{synch} module that uses suspension
instead of spinlocks.
Figure~\ref{fig:suspension} shows the same \texttt{Lock} interface implemented
using suspension.
A \texttt{Lock} maintains both a boolean indicating whether the lock is taken,
and a list of contexts of processes that want to acquire the lock.
\texttt{lock} sets the lock if available, or suspends itself if not.
Note that \texttt{stop} is called within an \texttt{atomic} statement---this is
the only exception to an atomic statement running to completion.  While the
process is running no other processes can run, but when it suspends itself
it allows other processes to run.

\texttt{unlock} checks to see if any processes are waiting to get the lock.
If so, it uses the \texttt{head} and \texttt{tail}
methods from the \texttt{list} module to resume the first process that got
suspend and to remove its context from the list.
Selecting the first process is a design choice.  Another implementation could
have picked the last one, and yet another implementation could have used
\texttt{choose} to pick an arbitrary one.  Selecting the first is a common
choice in lock implementations as it prevents \emph{starvation}: every process
gets a chance to obtain the lock (assuming every process eventually releases
it).  Chapter~\ref{ch:starvation} will talk more about starvation.

CXL allows you to select which version of the \texttt{synch} module you would
like to use with the \texttt{-m} flag.  For example,

\begin{code}
\begin{verbatim}
cxl -m synch=synchS x.cxl
\end{verbatim}
\end{code}

runs the file \texttt{x.cxl} using the Suspension version of the \texttt{synch} module.
You will find that using the \texttt{synchS} module often leads to searching a
significantly larger search space than using the \texttt{synch} module.
Part of the reason is that the \texttt{synchS} module keeps track of the order
in which processes wait for a lock.

@@

The astute reader might wonder if obtaining an ordinary lock itself is an
example of busy-waiting.  After all, the CXL \texttt{synch} implementation
of \texttt{lock()} spins in a loop until the lock is available
(see Figure~\ref{fig:spinlocks}).
In most operating systems and programming language runtimes, however,
when a process acquires a lock, the process is placed on a scheduling
queue and stops using CPU cycles until the lock becomes available.
Moreover, if there are multiple processes waiting for a lock, there
is a scheduler that decides which of the processes gets the lock.
Busy waiting disables all this.

@@

A process trying to obtain a read or write lock in
Figure~\ref{fig:rwbusy} has to obtain a spinlock just to check
if the read or write lock is available.  So even when the read or write
lock is not available, the process ends up updating the spinlock and
therefore the shared memory.

@@

\chapter{Starvation}
\label{ch:starvation}
